---
title: "RFC: Migration to PROJ6/GDAL3"
author: "Roger Bivand"
output:
  html_document:
    toc: true
bibliography: PROJ.bib
link-citations: yes
vignette: >
  %\VignetteIndexEntry{RFC: Migration to PROJ6/GDAL3}
  %\VignetteEngine{knitr::knitr}
  %\VignetteEncoding{UTF-8}
---

# Migration to PROJ6/GDAL3

### PROJ

Because so much open source (and other) software uses the PROJ library and framework, many are affected when PROJ upgrades. Until very recently, PROJ has been seen as very reliable, and the changes taking place now are intended to confirm and reinforce this reliability. Before PROJ 5 (PROJ 6 is out now, PROJ 7 is coming early in 2020), the `+datum=` tag was used, perhaps with `+towgs84=` with three or seven coefficients, and possibly `+nadgrids=` where datum transformation grids were available. However, transformations from one projection to another first inversed to longitude-latitude in WGS84, then projected on to the target projection.


### Big bump coming:

'Fast-forward 35 years and PROJ.4 is everywhere: It provides coordinate handling for almost every geospatial program, open or closed source. Today,  we  see  a  drastical  increase  in  the  need  for  high  accuracy  GNSS  coordinate  handling, especially in the agricultural and construction engineering sectors.  This need for geodetic-accuracy transformations  is  not  satisfied  by  "classic  PROJ.4".  But  with  the  ubiquity  of  PROJ.4,  we  can provide these transformations "everywhere", just by implementing them as part of PROJ.4' [@evers+knudsen17].


### Escaping the WGS84 hub/pivot: PROJ and OGC WKT2

Following the introduction of geodetic modules and pipelines in PROJ 5 [@knudsen+evers17; @evers+knudsen17], PROJ 6 moves further. Changes in the legacy PROJ representation and WGS84 transformation hub have been coordinated through the [GDAL barn raising](https://gdalbarn.com/) initiative. Crucially WGS84 often ceases to be the pivot for moving between datums. A new OGC WKT is coming, and an SQLite EPSG file database has replaced CSV files. SRS will begin to support 3D by default, adding time too as SRS change. See also [PROJ migration notes](https://proj.org/development/migration.html).

There are very useful postings on the PROJ mailing list from Martin Desruisseaux, first [proposing clarifications](https://lists.osgeo.org/pipermail/proj/2019-July/008748.html) and a [follow-up](https://lists.osgeo.org/pipermail/proj/2019-August/008750.html) including a summary:

> * "Early binding" ≈ hub transformation technique.

> * "Late binding" ≈ hub transformation technique NOT used, replaced by
a more complex technique consisting in searching parameters in the
EPSG database after the transformation context (source, target,
epoch, area of interest) is known.

> * The problem of hub transformation technique is independent of WGS84.
It is caused by the fact that transformations to/from the hub are
approximate. Any other hub we could invent in replacement of WGS84
will have the same problem, unless we can invent a hub for which
transformations are exact (I think that if such hub existed, we
would have already heard about it).

> The solution proposed by ISO 19111 (in my understanding) is:

> * Forget about hub (WGS84 or other), unless the simplicity of
early-binding is considered more important than accuracy.

> * Associating a CRS to a coordinate set (geometry or raster) is no
longer sufficient. A {CRS, epoch} tuple must be associated. ISO
19111 calls this tuple "Coordinate metadata". From a programmatic
API point of view, this means that getCoordinateReferenceSystem()
method in Geometry objects (for instance) needs to be replaced by a
getCoordinateMetadata() method.

### Impacts of GDAL barnraising on **sp** workflows

The following examples will contrast the behaviour of PROJ4/GDAL2 (similar to PROJ5/GDAL2, and PROJ4/GDAL1) and PROJ6+/GDAL3+. In particular, the behaviour of the `exportToProj4()` function in GDAL's `OGRSpatialReference` class has changed:

> Warning
    Use of this function is discouraged. Its behaviour in GDAL >= 3 / PROJ >= 6 is significantly different from earlier versions. In particular +datum will only encode WGS84, NAD27 and NAD83, and +towgs84/+nadgrids terms will be missing most of the time. PROJ strings to encode CRS should be considered as a a legacy solution. Using a AUTHORITY:CODE or WKT representation is the recommended way.
    
(https://gdal.org/doxygen/classOGRSpatialReference.html#a271b3de4caf844135b0c61e634860f2b); see also (https://github.com/r-spatial/sf/issues/1187) and links therein to (https://github.com/r-spatial/discuss/issues/28) and (https://github.com/r-spatial/sf/issues/1146).

This function is used for both raster and vector data read through GDAL to provide the PROJ 4 string used to specify the coordinate reference system of **sp** `"Spatial"` objects using the `"CRS"` (S4, new style) class. Such classes cannot be modified without making it impossible for users to load serialised objects, such as **sp** RDS objects from GADM [for example for Norway](https://gadm.org/download_country_v3.html).

My fork of **sp** (https://github.com/rsbivand/sp) is currently at 1.3-3 or higher, and contains extra code conditionally using the **rgdal** R-Forge repository at 1.5-1 or higher. The commented out blocks were generated on a Windows platform, using PROJ4/GDAL2.


```{r}
library(sp)
packageVersion("sp")
```

```{r}
## [1] '1.3.2'
```

The `"CRS"` class definition is unchanged going forward to maintain backward compatibility.

```{r}
getClass("CRS")
```

```{r}
rgdal::rgdal_extSoftVersion()
```

```{r}
##           GDAL GDAL_with_GEOS         PROJ.4             sp 
##        "2.2.3"         "TRUE"        "4.9.3"        "1.3-1"
```

```{r}
packageVersion("rgdal")
```

```{r}
## [1] '1.4.7'
```

The changes introduced affect `CRS()` when **rgdal** is available; if **rgdal** is not available, it is just a lightly checked string. Because some terms are deprecated or defunct, we need to be careful. GDAL's `exportToProj4()` uses the PROJ `proj_as_proj_string()` function in its new API to return the PROJ string. Terms which are deprecated or defunct may be omitted. In the PROJ6+/GDAL3+ case, `CRS()` calls `rgdal::checkCRSArgs_ng()`, calling `RGDAL_checkCRSArgs()`, a compiled function

```{r}
(crs <- CRS("+proj=longlat +ellps=WGS84"))
```

```{r}
comment(crs)
```

For the PROJ4/GDAL2 (and similar) cases, if **rgdal** is available, `CRS()` calls `rgdal::checkCRSArgs()`, calling `RGDAL_checkCRSArgs()`, a compiled function, which calls `pj_init_plus()` in PROJ to check the validity of the string and possibly expand terms. If this succeeds, `pj_get_def()` is used to return the PROJ string. Both of these functions are part of the deprecated PROJ API that is still accessible in PROJ 6, but will soon be made defunct. 

```{r}
## CRS arguments: +proj=longlat +ellps=WGS84
```


```{r}
(crs <- CRS("+proj=longlat +datum=WGS84"))
```

```{r}
comment(crs)
```

```{r}
## CRS arguments:
##  +proj=longlat +datum=WGS84 +ellps=WGS84 +towgs84=0,0,0
```


```{r}
(crs <- CRS("+proj=longlat +towgs84=0,0,0"))
```

```{r}
comment(crs)
```

```{r}
## CRS arguments:
##  +proj=longlat +towgs84=0,0,0 +ellps=WGS84
```

```{r}
(crs <- CRS("+init=epsg:4326"))
```

```{r}
comment(crs)
```

```{r}
## CRS arguments:
##  +init=epsg:4326 +proj=longlat +datum=WGS84 +no_defs +ellps=WGS84
## +towgs84=0,0,0
```


```{r}
(crs <- CRS("+init=epsg:27700"))
```

```{r}
comment(crs)
```

```{r}
## CRS arguments:
##  +init=epsg:27700 +proj=tmerc +lat_0=49 +lon_0=-2 +k=0.9996012717
## +x_0=400000 +y_0=-100000 +datum=OSGB36 +units=m +no_defs
## +ellps=airy
## +towgs84=446.448,-125.157,542.060,0.1502,0.2470,0.8421,-20.4894
```



```{r}
(crs <- CRS(SRS_string = "EPSG:27700"))
```

```{r}
comment(crs)
```

```{r}
(crs <- CRS("+proj=tmerc +lat_0=49 +lon_0=-2 +k=0.9996012717 +x_0=400000 +y_0=-100000 +datum=OSGB36 +units=m +no_defs"))
```


```{r}
comment(crs)
```


```{r}
cat(rgdal::showSRID("+init=epsg:27700", format="WKT2_2018", multiline="YES"), "\n")
```

